---
title: "Student Surveys"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Student Surveys}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(knitr)
library(tntpmetrics)
```

`tntpmetrics` calculates engagement, relevance, and belonging metrics from student surveys. The list below shows the required column names to calculate each metric.

* **Engagement**
    * Metric name to use in package: `metric = engagement`
    * Items: ("*We spend time in class on things that...*")
      + eng_interest ("*What we were learning was interesting.*")
      + eng_like ("*I liked what we did in class.*")
      + eng_losttrack ("*I was so into what we were learning I lost track of time.*")
      + eng_moreabout ("*I thought more about what we were learning than anything else.*")
    * Scale: 0 (Not True), 1 (A Little True), 2 (Mostly True), or 3 (Very True).
* **Relevance**
    * Metric name to use in package: `metric = relevance`
    * Items: ("*We spend time in class on things that...*")
      + rel_asmuch ("*Will help me learn just as much as kids in other schools.*")
      + rel_future ("*Are important to my future goals.*")
      + rel_outside ("*I can use outside of school.*")
      + rel_rightnow ("*Are important to my life right now.*")
    * Scale: 0 (Not True), 1 (A Little True), 2 (Mostly True), or 3 (Very True).
* **Belonging**
    * Metric name to use in package: `metric = belonging`
    * Items:
      + bel_ideas ("*In this class, my ideas really count.*")
      + tch_interestedideas ("*In this class, my teacher is interested in my ideas.*")
      + bel_fitin ("*In this class, I feel like I fit in.*")
      + tch_problem ("*I could talk to my teacher for this class if I had a problem.*")
    * Scale: 0 (Not True), 1 (A Little True), 2 (Mostly True), or 3 (Very True).

## Engagement

```{r}
# data set in tntpmetrics package
data(ss_data_initial)

ss_data_initial %>%
  select(response_id, starts_with('eng')) %>%
  make_metric(metric = "engagement") %>%
  head() %>%
  kable()
```

## Relevance

```{r}
ss_data_initial %>%
  select(response_id, starts_with('rel')) %>%
  make_metric(metric = "relevance") %>%
  head() %>%
  kable()
```

## Belonging

```{r}
ss_data_initial %>%
  select(response_id, starts_with('bel'), starts_with('tch')) %>%
  make_metric(metric = "belonging") %>%
  head() %>%
  kable()
```
